---
title: "Logistic Regression Model"
author: "SQR"
date: "2024-08-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
alz_data <- read.csv('alzheimers_disease_data.csv')
alz_data <- alz_data[,-c(1,35)] 
```

```{r}
alz_cat_data = alz_data %>%
  mutate(Gender = ifelse(Gender == 0,"M","F")) %>%
  mutate(Ethnicity = ifelse(Ethnicity == 0, "Caucasian",
                     ifelse(Ethnicity == 1, "African American",
                     ifelse(Ethnicity == 2, "Asian", "Other")))) %>%
  mutate(EducationLevel = ifelse(EducationLevel == 0, "None",
                          ifelse(EducationLevel == 1, "High School",
                          ifelse(EducationLevel == 2, "Bachelor's", "Higher")))) %>%
  mutate(Smoking = ifelse(Smoking == 0, "No", "Yes")) %>%
  mutate(FamilyHistoryAlzheimers = ifelse(FamilyHistoryAlzheimers == 0, "No", "Yes")) %>%
  mutate(CardiovascularDisease = ifelse(CardiovascularDisease == 0, "No", "Yes")) %>%
  mutate(Diabetes = ifelse(Diabetes == 0, "No", "Yes")) %>%
  mutate(Depression = ifelse(Depression == 0, "No", "Yes")) %>%
  mutate(HeadInjury = ifelse(HeadInjury == 0, "No", "Yes")) %>%
  mutate(Hypertension = ifelse(Hypertension == 0, "No", "Yes")) %>%
  mutate(MemoryComplaints = ifelse(MemoryComplaints == 0, "No", "Yes")) %>%
  mutate(BehavioralProblems = ifelse(BehavioralProblems == 0, "No", "Yes")) %>%
  mutate(Confusion = ifelse(Confusion == 0, "No", "Yes")) %>%
  mutate(Disorientation = ifelse(Disorientation == 0, "No", "Yes")) %>%
  mutate(PersonalityChanges = ifelse(PersonalityChanges == 0, "No", "Yes")) %>%
  mutate(DifficultyCompletingTasks = ifelse(DifficultyCompletingTasks == 0, "No", "Yes")) %>%
  mutate(Forgetfulness = ifelse(Forgetfulness == 0, "No", "Yes"))
```


Reconstruct the corrplot with MMSE included this time:
```{r}
numerics <- alz_data%>%
  dplyr::select(Age, BMI, AlcoholConsumption, PhysicalActivity, DietQuality, SleepQuality, SystolicBP,
                DiastolicBP, CholesterolTotal, CholesterolLDL, CholesterolHDL, CholesterolTriglycerides,
                MMSE, FunctionalAssessment, ADL)

cor_mat = round(cor(numerics),2)
ggcorrplot::ggcorrplot(cor_mat, lab=TRUE, lab_size=2, tl.cex = 10, method = "square")+
  scale_fill_gradient2(limits = c(-0.2, 0.2), low = "blue", mid = "white", high = "red", midpoint = 0, space = "Lab", name = "Correlation")

```
No difference. 


## Construct the logit model

First with all predictors, no variable transformations:
```{r}
mod1 <- glm(Diagnosis~., alz_cat_data, family='binomial')
summary(mod1)
```

Reduce model to only include significant predictors; here we will only include predictors that are significant at the 0.05 level:
```{r}
mod2 <- glm(Diagnosis~CholesterolLDL+MMSE+FunctionalAssessment+MemoryComplaints+ADL, alz_data, family='binomial')
summary(mod2)
```

Now using AIC to select variables:
```{r}
aic = MASS::stepAIC(mod1, direction='both', trace=F)
summary(aic)
```


## Incorporate cross validation
```{r}
library(caret)
alz_cat_data <- alz_cat_data%>%
  mutate(Diagnosis_cat = ifelse(Diagnosis==1,'Yes','No'))
ctrl <- trainControl(method='cv',
                            number=10,
                            summaryFunction=twoClassSummary,
                            classProbs=TRUE,
                            savePredictions=TRUE)

mod_cv <- train(Diagnosis_cat~.-Diagnosis, data=alz_cat_data, method='glmStepAIC', family='binomial',
                trControl=ctrl, metric='ROC', trace=FALSE)
summary(mod_cv)
```


## Assess model performance
```{r}
predictions <- mod_cv$pred
```

```{r}
table(predictions$pred, predictions$obs) # cross validated confusion matrix
```

```{r}
library(pROC)

my_roc <- roc(predictions$obs, predictions$Yes)

roc_dat <- data.frame(TPR=my_roc$sensitivities, FPR=1-my_roc$specificities)
ggplot(roc_dat, aes(x=FPR, y=TPR))+geom_line(color='lightpink')+theme_minimal()
```

## Linearity assumption
```{r}
alz_cat_data <- alz_cat_data%>%
  mutate(prob=predict(mod_cv$finalModel, type='response'), log_odds=log(prob/(1-prob)))
```

```{r}
# Get predictors in final model
mod_cv$finalModel
```


```{r}
# Scatter plots of log odds vs numerical predictors
long <- gather(alz_cat_data, key="predictor", value ="value", 
                   Age, SleepQuality, CholesterolLDL, CholesterolHDL, MMSE, FunctionalAssessment, ADL)

ggplot(long, aes(x=value, y=log_odds, color=predictor))+
  geom_point(size=0.01)+
  facet_wrap(~predictor, scales = "free_x", ncol=4)+theme_minimal()
```


